Fundamental Problems with the Evidence Base of Adolescent Depression Treatments.

### 2024-03-06 apples_and_oranges.qmd
Cleaning control dataset, few other changes to code

# # We also need to calculate SE for proportion women for the baseline calculations
# # for proportions, this is calculated as sqrt(p(1-p)/n), which I implement stepwise below

@@

The data for the studies included in this metanalysis are summarised in Supplementary Table 1 and are also available as a csv dataframe on \[<https://github.com/transatlantic-comppsych/apples_oranges>\]. <!--# Charlotte, this should essentially be our df_long dataset, right? Perhaps, for the puproses of the table we can get rid of some of the columns. It would be good to have it in the way Pim had it. CB - I'm not sure I entirely understand. df_long_for_metan is already quite succinct (6 columns). I wonder if you are thinking of df_appl_v_orange which is more complete and includes means and sds etc. I'd say that's probably more appropriate to make available. I agree we should simplify it down a bit. Happy to do that once you confirm I'm on the right track here. If we decide to format it to be consistent with Cuijpers' data, we would have one row per comparison, rather than the long format we have in df_long_for_metan. -->

@@

Gender composition was similar across psychotherapy and medication trials,

@@

#### Quality and nature of psychotherapy controls across trials

@@

>>>>

# We also need to calculate SE for proportion women for the baseline calculations
# for proportions, this is calculated as sqrt(p(1-p)/n), which I implement stepwise below

@@

The data for the studies included in this metanalysis are summarised in Supplementary Table 1 and are also available as a csv dataframe on \[<https://github.com/transatlantic-comppsych/apples_oranges>\].

@@

Gender composition did not differ significantly between psychotherapy and medication trials (stats).

@@

@@

### Comparing the nature and intensity of control conditions in psychotherapy trials

Text.

# We need to read in the appropriate dataset which describes the active and control conditions used in
# psychotherapy trials.

df_descr_psy_conditions <- readxl:: read_excel("for_despina_and_giannis_200224.xlsx")

# checking id matchup
# discrepancies <- setdiff(df_descr_psy_conditions$new_study_id, df_long_for_metan$new_study_id)

# Cleaning the dataset
# For now lets look at variables pertaining to the nature / intensity of interventions

colnames(df_descr_psy_conditions)

df_descr_psy_conditions <- df_descr_psy_conditions %>% 
  select(c(new_study_id, `Psychotherapy/control`, levels_var, number_of_sites, Number_of_sessions,
           frequency_weeks, Length_of_sessions_mins, Total_hours_of_intervention_hours, total_hours_gsh,
           Total_period_of_intervention_weeks, Format_group_individual, Delivery, Involvement_of_others))

# rename for clarity and to match master dataframe
df_descr_psy_conditions <- df_descr_psy_conditions %>% 
  rename(treatment = 'Psychotherapy/control', no_sites = number_of_sites, no_sessions =
           Number_of_sessions, freq_weeks = frequency_weeks, length_sessions_mins =
           Length_of_sessions_mins, total_hours = Total_hours_of_intervention_hours,
         total_period_weeks = Total_period_of_intervention_weeks, format = Format_group_individual, delivery =
           Delivery, family_involv = Involvement_of_others)

check <- df_descr_psy_conditions %>% 
  filter(treatment == 'wlc')

# Visual inspection

# There are two studies where the waitlist condition has some intervention (i.e. values should not be set to 0)
# Otherwise, I want values to be set to 0, not NA.

df_descr_psy_conditions <- df_descr_psy_conditions %>%
  mutate(
    no_sessions = ifelse(treatment == "wlc" & !(new_study_id %in% c("Ackerson, 1998", "Diamond, 2002")), 0, no_sessions),
    freq_weeks = ifelse(treatment == "wlc" & !(new_study_id %in% c("Ackerson, 1998", "Diamond, 2002")), 0, freq_weeks),
    length_sessions_mins = ifelse(treatment == "wlc" & !(new_study_id %in% c("Ackerson, 1998", "Diamond, 2002")), 0, length_sessions_mins),
    total_hours = ifelse(treatment == "wlc" & !(new_study_id %in% c("Ackerson, 1998", "Diamond, 2002")), 0, total_hours)
  )

# Now cau

check <- df_descr_psy_conditions %>% 
  filter(treatment == 'cau')

# All studies except for Martinovic, Srivastava, Stikkelbroek and Weisz 2009 should have variables set to NA. 
# This is because we cannot say that they did not feature 0 hours of intervention, but that for the most part
# it is not reported or it is not possible to quantify.

df_descr_psy_conditions <- df_descr_psy_conditions %>%
  mutate(
    no_sessions = ifelse(treatment == "cau" & !(new_study_id %in% c("Martinovic, 2006", "Srivastava, 2020", "Stikkelbroek, 2020", "Weisz, 2009")), NA, no_sessions),
    freq_weeks = ifelse(treatment == "cau" & !(new_study_id %in% c("Martinovic, 2006", "Srivastava, 2020", "Stikkelbroek, 2020", "Weisz, 2009")), NA, freq_weeks),
    length_sessions_mins = ifelse(treatment == "cau" & !(new_study_id %in% c("Martinovic, 2006", "Srivastava, 2020", "Stikkelbroek, 2020", "Weisz, 2009")), NA, length_sessions_mins),
    total_hours = ifelse(treatment == "cau" & !(new_study_id %in% c("Martinovic, 2006", "Srivastava, 2020", "Stikkelbroek, 2020", "Weisz, 2009")), NA, total_hours)
  )

# Now other controls

check <- df_descr_psy_conditions %>%
  filter(!(treatment %in% c('cau', 'wlc'))) %>% 
  filter(levels_var == "psychotherapy_control")

# Made a few manual adjustments in the original excel sheet.

# List of columns to clean
columns_to_clean <- c(
  "no_sessions",
  "freq_weeks",
  "length_sessions_mins",
  "total_hours",
  "total_period_weeks"
)

# Convert non-numeric values to NA in specified columns
df_descr_psy_conditions <- df_descr_psy_conditions %>%
  mutate_at(vars(columns_to_clean), ~as.numeric(as.character(.)))

# Define the function for summary stats

group_function <- function(dataframe, my_groups, columns) {
  result_dataframe <- data.frame(
    Group = character(),
    Column = character(),
    N = numeric(),
    Mean = numeric(),
    Std_Dev = numeric(),
    Median = numeric(),
    IQR = numeric(),
    stringsAsFactors = FALSE
  )
  unique_groups <- unique(dataframe[[my_groups]])
  for (group in unique_groups) {
    group_data <- dataframe[dataframe[[my_groups]] == group, ]
    for (col in columns) {
      col_data <- group_data[[col]]
      n <- sum(!is.na(col_data))
      mean_val <- mean(col_data, na.rm = TRUE)
      std_dev <- sd(col_data, na.rm = TRUE)
      median_val <- median(col_data, na.rm = TRUE)
      iqr <- IQR(col_data, na.rm = TRUE)
      result_dataframe <- rbind(result_dataframe,
                                data.frame(Group = group,
                                           Column = col, N = n, Mean = mean_val,
                                           Std_Dev = std_dev, Median = median_val,
                                           IQR = iqr, stringsAsFactors = FALSE))
    }
  }
  return(result_dataframe)
}

sessions <- group_function(df_descr_psy_conditions,"levels_var","no_sessions")
frequency <- group_function(df_descr_psy_conditions,"levels_var","freq_weeks")
length_sessions_mins <- group_function(df_descr_psy_conditions,"levels_var","length_sessions_mins")
total_hours <- group_function(df_descr_psy_conditions,"levels_var","total_hours")
total_period_weeks <- group_function(df_descr_psy_conditions,"levels_var","total_period_weeks")

control_summary <- rbind(sessions, frequency, length_sessions_mins, total_hours, total_period_weeks)
